---
title: "Data Input/Output"
author: "Steve Bellan (slides borrowed from John Muschelli)"
output:
  ioslides_presentation:
    css: ../styles.css
    widescreen: yes
---

```{r, echo = FALSE}
library(knitr)
library(readr)
opts_chunk$set(comment = "")
```
## Common new user mistakes we have seen

1.  **Working directory problems: trying to read files that R "can't find"**
    - RStudio can help, and so do RStudio Projects
    - discuss in Data Input/Output lecture
2.  Typos (R is **case sensitive**, `x` and `X` are different)
    - RStudio helps with "tab completion"
    - discussed throughout
3.  Data type problems (is that a string or a number?)
4.  Open ended quotes, parentheses, and brackets     
5.  Different versions of software
    
## Working Directories

* R "looks" for files on your computer relative to the "working" directory
* Many people recommend not setting a directory in the scripts 
    - assume you're in the directory the script is in
    - If you open an R file with a new RStudio session, it does this for you.
* If you do set a working directory, do it at the beginning of your script. 
* Example of getting and setting the working directory:

```{r workingDirectory,eval=FALSE}
## get the working directory
getwd()
setwd('~/Documents/R Repos/EPID7500/')
```


## Setting a Working Directory

* Setting the directory can sometimes be finicky
    * **Windows**: Default directory structure involves single backslashes ("\"), but R interprets these as "escape" characters. So you must replace the backslash with forward slashes ("/") or two backslashes ("\\")
    * **Mac/Linux**: Default is forward slashes, so you are okay
* Typical directory structure syntax applies
    * ".." - goes up one level
    * "./" - is the current directory
    * "~" - is your "home" directory

## Working Directory 

Note that the `dir()` function interfaces with your operating system and can show you which files are in your current working directory. 

You can try some directory navigation: 

```{r directoryNav}
dir() # shows directory contents
dir("..") # shows up one directory
dir("../..") # shows up two directories
```

## Relative vs. absolute paths (From Wiki)

An **absolute or full path** points to the same location in a file system, regardless of the current working directory.
```{r}
dir('~/Documents', full.names = TRUE)[1:3]
```

This means if I try your code, and you use absolute paths, it won't work unless we have the exact same folder structure where R is looking (bad).

A **relative path starts from current  working directory**. Need to have the working directory right but good for project portability.
```{r}
dir('~/Documents', full.names = FALSE)[1:3]
```

## Setting the Working Directory

In RStudio, go to `Session --> Set Working Directory --> To Source File Location`

RStudio should put code in the Console, similar to this:
```{r, eval = FALSE}
setwd("~/Lectures/Data_IO/lecture")
```
    
## Setting the Working Directory
    
Again, if you open an R file with a new RStudio session, it does this for you.  You may need to make this a default.

1. Make sure RStudio is the default application to open .R files
    * Mac - right click --> Get Info --> Open With: RStudio --> Change All
    * [Windows 7](https://support.microsoft.com/en-us/help/18539/windows-7-change-default-programs) and [Windows 10](https://www.cnet.com/how-to/how-to-set-default-programs-in-windows-10/)


## Reminder: Use R Help!

For any function, you can write `?FUNCTION_NAME`, or `help("FUNCTION_NAME")` to look at the help file:

```{r, eval = FALSE}
?dir
help("dir")
```

## Data Input

* Everything we do in class will be using real publicly available data - there are few 'toy' example datasets and 'simulated' data

* 'Reading in' data is the first step of any real project/analysis
* R can read almost any file format
* Focus on
    * tab delimited (e.g. '.txt')
    * comma separated (e.g. '.csv')
    * Microsoft Excel (e.g. '.xlsx')

## Data Input

[Youth Tobacco Survey (YTS) Dataset](https://catalog.data.gov/dataset/youth-tobacco-survey-yts-data)

* Middle/high school students  tobacco use, exposure to environmental tobacco smoke, smoking cessation, school curriculum,  knowledge and attitudes about tobacco, etc...

* Download the CSV file from the above site by clicking Download 
    * If a file loads in your browser, choose File --> Save As
* Within RStudio: Session --> Set Working Directory --> To Source File Location

## Data Input

R Studio features

* "File --> Import Dataset --> From CSV" command.
* Specify the formatting of your text file. 
* Shows you the corresponding R commands that did the import

## Read in from the internet directly or from a downloaded copy

* Web

```{r read_url, message = FALSE}
mydat = read_csv("https://github.com/sbellan61/EPID7500-IntroToR/raw/gh-pages/1.3-Data_IO/Youth_Tobacco_Survey_YTS_Data.csv")
head(mydat)
```

* File

```{r readCSV2}
dat = read_csv("../data/Youth_Tobacco_Survey_YTS_Data.csv")
```

The data is now successfully read into your R workspace, just like from using the dropdown menu.

## Base R vs tidyverse

* Base R (loads as a data.frame): read.csv, read.table, read.delim
* TidyVerse (loads as a tibble): read_csv, read_table, read_delim

* [tibbles](http://r4ds.had.co.nz/tibbles.html) print cleaner and have slightly different subsetting syntax

## Data Input

```{r, viewTibble}
dat
```

While many online resources use the base R tools, the latest version of RStudio switched to use these new `readr` data import tools, so we will use them in the class for slides. They are also up to two times faster for reading in large datasets, and have a progress bar which is nice. 


## Data Input

Here is how to read in the same dataset using base R functionality, which returns a `data.frame` directly

```{r, readingCSV}
dat2 = read.csv("../data/Youth_Tobacco_Survey_YTS_Data.csv", as.is = TRUE)
head(dat2)
```

We will use the TidyVerse `readr` functions because TidyVerse is the wave of the future.

## Data Input

* `nrow()` displays the number of rows of a data frame
* `ncol()` displays the number of columns 
* `dim()` displays a vector of length 2: # rows, # columns
* `colnames()` displays the column names (if any) and `rownames()` displays the row names (if any)

```{r, dims}
dim(dat2)
nrow(dat2)
ncol(dat2)
colnames(dat2)
```


## Data Input

Changing variable names in `data.frame`s works using the `names()` function, which is analagous to `colnames()` for data frames (they can be used interchangeably)

```{r names1}
names(dat)[1] = "year"
names(dat)
```


## Data Output

While its nice to be able to read in a variety of data formats, it's equally important to be able to output data somewhere.

There are also data exporting functions in the `readr` package, which have the pattern `write_*` like `write_csv` and `write_delim`

```
write_csv(x, path, na = "NA", append = FALSE, col_names = !append)
```

## Data Output

`x`: the R `data.frame` or `matrix` you want to write

`path`: the file name where you want to R object written. It can be an absolute path, or a filename (which writes the file to your working directory)

* you can also control other things like whether to print rownames (check out the helpfile!)

## Data Output

For example, we can write back out the Youth Tobacco dataset with the new column name:

```{r writecsv,eval=FALSE}
names(dat)[1] = "Year"
write_csv(dat, path="YouthTobacco_newNames.csv")
```

## Data Input - Excel

Excel to R options

* Saving the Excel sheet as a .csv file, and using `read_csv()`
* Using an add-on package, like `xlsx`, `readxl`, or `openxlsx`

For single worksheet .xlsx files, I often just save the spreadsheet as a .csv file (because I often have to strip off additional summary data from the columns)

For an .xlsx file with multiple well-formated worksheets, I use the `readxl` package for reading in the data. 

## Data Input - Other Software

* **haven** package (https://cran.r-project.org/web/packages/haven/index.html) reads in SAS, SPSS, Stata formats
* **readxl** package - the `read_excel` function can read Excel sheets easily
* **readr** package - Has *read_csv*/*write_csv* and *read_table* functions similar to *read.csv*/*write.csv* and *read.table*.  Has different defaults, but can read **much faster** for very large data sets
* **sas7bdat** reads .sas7bdat files
* **foreign** package - can read all the formats as **haven**.  Around longer (aka more testing), but not as maintained (bad for future).

## More ways to save: save

The `save` command can save a set of R objects into an "R data file", with the extension `.rda` or `.RData`.  

```{r, message = FALSE}
x = 5; # can have semicolons a the end!
# calling read_csv function and pasting a long string together
yts = readr::read_csv(
  paste0("http://johnmuschelli.com/intro_to_r/",
         "data/Youth_Tobacco_Survey_YTS_Data.csv")) 
save(yts, x, file = "yts_data.rda")
```

## More ways to save: load

The opposite of `save` is `load`.  The `ls()` command lists the items in the workspace/environment and `rm` removes them:

```{r}
ls() # list things in the workspace
rm(list = c("x", "yts"))
ls()
z = load("yts_data.rda")
ls()
```

## More ways to save: load

```{r}
print(z)
```

Note, `z` is a **character vector** of the **names** of the objects loaded, **not** the objects themselves.
