---
title: "Title"
date: Today
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the necessary packages

Note that to enter a code segment we can do `Ctrl-Alt-i` (Windows) or  `Cmd-Alt-i` (Mac)
```{r message=F}
require(tidyverse)
```

## Knitting

* Use keyboard shortcuts to knit whenever you want
* `Ctrl-Shift-k` (Windows) or `Cmd-Shift-k` (Mac)

## Loading in a data set from the internet

* [WHO TB notification data provided by countries](http://www.who.int/tb/country/data/download/en/)
* Right click on `Download case notifications [1.9Mb]` and save it to your hard drive
* Cannot load directly from internet since no URL for the CSV file available
* [WHO Data Dictionary](https://extranet.who.int/tme/generateCSV.asp?ds=dictionary)


## Workflow for running R code within Rmd

* I prefer to be able to use Console
* Set R output to console (not inline) when running lines [see here](https://support.rstudio.com/hc/en-us/community/posts/115000797947-Global-Changes-show-chunk-output-in-console-not-execute-all-lines)

## Select just a few variables of interest


## Missing values


* Excercise: Repeat this but look at rows with 0 values

## Identify countries without missing rows

## Subsetting

* Select the data set for which there are no NAs

## Excercises

1. Write code that spreads `notif3` so that each country has it's own column
1. Look at a boxplot of TB notifications by country for the 10 countries with the greatest # of notifications overall.
    + Hint: `group_by()` countries, then sum all notifications, then sort 
    in descending order with `?arrange`

## Answers 1

1. Write code that spreads `notif3` so that each country has it's own column

## Answers 2

2. Look at a boxplot of TB notifications by country for the 10 countries with the greatest # of cumulative notifications across years.

## Answers 2

2. Look at a boxplot of TB notifications by country for the 10 countries with the greatest # of cumulative notifications across years.

## Answers 2 (with ggplot)

2. Look at a boxplot of TB notifications by country for the 10 countries with the greatest # of cumulative notifications across years.


## Excercise 3

* [WHO Measles & Rubella Page](http://www.who.int/immunization/monitoring_surveillance/burden/vpd/surveillance_type/active/measles_monthlydata/en/)
* Download the [Distribution of measles cases by country and by month, 2011-2017](http://www.who.int/entity/immunization/monitoring_surveillance/burden/vpd/Table_407_Web_Number_Measles_Cases_Month_2017-08-09.xlsx?ua=1)
* Read it into R with the  `xlsx` library (note you have to read 2nd sheet!)
* Reformat the data so it is in long with date in YYYY-MM format

## Answer 3

## Excercise 4

* Plot measles incidence time series by country for the 5 highest incidence countries in 2015

## Export Rmd to R

* You can extract the inside of code chunks using `purl()`

